---
title: "Bayesian Loss Function - Operating Characteristics"
geometry: margin=0.5in
---

The following output closely follows the output given within the Shiny app. 
Each subheader here contains output from the corresponding tab.  

\  

# Simulation Results

```{r simrestable, fig.width=8, echo=FALSE, comment=NA}
print(sim_res(), row.names=FALSE)
```

# Power Curve

```{r powercurves, fig.width=8, fig.height=6, echo=FALSE}
p3()
```


# Quantile plot

```{r quantileplots, fig.width=8, fig.height=6, echo=FALSE}
pp3()
```

\  

# Input values
The following values were used for the simulation

* Simulation inputs
    + Distribution: `r input$distribution`
    + True rate: `r inputVec()$true_rate`
    + Weibull scale: `r inputVec()$weibull_scale`
    + Weibull scale: `r inputVec()$weibull_shape`
    + Number of simulations: `r input$sim`
* Study inputs
    + Confidence level: `r input$CL`
    + Number of current subjects: `r input$N`
    + Number of historical subjects: `r input$N0`
    + Number of historical events: `r input$y0`
    + Null rate: `r input$H0_rate`
    + Number of MCMC simulations: `r input$number_mcmc`
* Prior settings 
    + Max effective sample size: `r input$N0_max`
    + alpha - initial non-informative prior: `r input$alpha0`
    + beta - initial non-informative prior: `r input$beta0`
    